---
title: "Cyclistic Bike Share"
author: "Cipriano Arralde"
date: "2024-06-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setting my environment
```{r Uploading libraries}
library(tidyverse)
library(ggplot2)
library(lubridate)
library(janitor)
library(tidyr)
library(readr)
library(skimr)
library(dplyr)
```

## Uploading the data

Downloaded the 12 month of 2021 from the link provided by the case study guide, [link here](https://divvy-tripdata.s3.amazonaws.com/index.html)

#### Loading up my files
```{r Uploading files, message=FALSE, warning=FALSE}
jan2021 <- read.csv("C:/Users/Cipriano/Documents/2021/202101-divvy-tripdata.csv")
feb2021 <- read.csv("C:/Users/Cipriano/Documents/2021/202102-divvy-tripdata.csv")
mar2021 <- read.csv("C:/Users/Cipriano/Documents/2021/202103-divvy-tripdata.csv")
apr2021 <- read.csv("C:/Users/Cipriano/Documents/2021/202104-divvy-tripdata.csv")
may2021 <- read.csv("C:/Users/Cipriano/Documents/2021/202105-divvy-tripdata.csv")
jun2021 <- read.csv("C:/Users/Cipriano/Documents/2021/202106-divvy-tripdata.csv")
jul2021 <- read.csv("C:/Users/Cipriano/Documents/2021/202107-divvy-tripdata.csv")
aug2021 <- read.csv("C:/Users/Cipriano/Documents/2021/202108-divvy-tripdata.csv")
sep2021 <- read.csv("C:/Users/Cipriano/Documents/2021/202109-divvy-tripdata.csv")
oct2021 <- read.csv("C:/Users/Cipriano/Documents/2021/202110-divvy-tripdata.csv")
nov2021 <- read.csv("C:/Users/Cipriano/Documents/2021/202111-divvy-tripdata.csv")
dec2021 <- read.csv("C:/Users/Cipriano/Documents/2021/202112-divvy-tripdata.csv")
```

## Creating the data frames

#### Combined all 12 files into one data fram

```{r Created initial data frame, message=FALSE, warning=FALSE}
cyclistic_2021_trips <- rbind(jan2021, feb2021, mar2021, apr2021, may2021, jun2021, jul2021, aug2021, sep2021, oct2021, nov2021, dec2021)
```

#### Selected columns for analysis

```{r Trimmed data frame, message=FALSE, warning=FALSE}
trimmed_cyclistic <- select(cyclistic_2021_trips, "rideable_type", "started_at", "ended_at", "member_casual")
```

#### Created second data frame for calculations

```{r Second data frame, message=FALSE, warning=FALSE}
final_cyclistic <- trimmed_cyclistic
```

## Cleaning

#### Created new column for the duration of the ride, then rounded the result to 1 decimal

```{r Trip duration column, message=FALSE, warning=FALSE}
final_cyclistic$ride_length <- difftime(final_cyclistic$ended_at, final_cyclistic$started_at, units = "mins") # Ride length column
final_cyclistic$ride_length <- round(final_cyclistic$ride_length, digits = 1) # Rounded to 1 decimal
```

#### Created date column with datetime format

```{r Date column, message=FALSE, warning=FALSE}
final_cyclistic$date <- as.POSIXlt(final_cyclistic$started_at, format = "%Y-%m-%d %H:%M") # Date column, datetime format
```

#### Exctracted the day of the week

```{r Weekday column, message=FALSE, warning=FALSE}
final_cyclistic$weekday <- weekdays(final_cyclistic$date) # Weekday column
```

#### Extracted the month

```{r Month column, message=FALSE, warning=FALSE}
final_cyclistic$month <- format(as.Date(final_cyclistic$date), "%m") # Month column
```

#### Extracted the year

```{r Year column, message=FALSE, warning=FALSE}
final_cyclistic$year <- format(as.Date(final_cyclistic$date), "%Y") # Year column
```

#### Extracted the day number

```{r Day column, message=FALSE, warning=FALSE}
final_cyclistic$day <- format(as.Date(final_cyclistic$date), "%d") # Day column
```

#### Extracted the hour

```{r Hour column, message=FALSE, warning=FALSE}
final_cyclistic$hour <- hour(final_cyclistic$date) # Hour column
```

#### Discarded duplicates and blanks

```{r Removed blanks and duplicates, message=FALSE, warning=FALSE}
final_cyclistic <- na.omit(final_cyclistic) # Removing blanks
final_cyclistic <- distinct(final_cyclistic) # Removing duplicates
```


#### Discarded values with trip duration <= 0

```{r Trip duration less or equal than 0, message=FALSE, warning=FALSE}
cyclistic_2021_df <- final_cyclistic[!(final_cyclistic$ride_length <= 0),] # New data frame, discarded rows where trip_duration <= 0
```

#### Renamed columns

```{r Renamed columns, message=FALSE, warning=FALSE}
cyclistic_2021_df <- cyclistic_2021_df %>% 
                      rename(bike_type = rideable_type,
                             trip_duration = ride_length,
                             user_type = member_casual) # Renaming some of the columns for personal preference
```

#### Created last data frame

```{r Last data frame, message=FALSE, warning=FALSE}
cyclistic_final_df <- select(cyclistic_2021_df, "hour",
                             "weekday", "day", "month", "year",
                             "bike_type", "user_type", "trip_duration") # Last data frame, cleaned and organized
```

## Quick Statistics

- **5.586.693** total rides
- **22 min** general average trip duration
- **Summer** busiest season
- **2.526.236 (45%)** total rides for **casual** users
- **3.060.457 (55%)** total rides for **member** users
- **32 min** average trip duration for **casual** users
- **13 min** average trip duration for **member** users
- **Saturday** busiest day for **casual** users
- **Wednesday** busiest day for **member** users
- **July** busiest month for **casual** users
- **September** busiest month for **member** users

## Vizzes

#### Total Trips By User Type

```{r Total Trips By User Type}
cyclistic_final_df %>% 
  group_by(user_type) %>% 
  summarize(num_of_rides = n()) %>%
  arrange(user_type) %>% 
  ggplot(aes(x = user_type, y = num_of_rides, fill = user_type)) +
  labs(title = "Total Trips By User Type") +
  geom_col(width = 0.5, position = position_dodge(width = 0.5)) + 
  scale_y_continuous(labels = function(x) format(x, scientific = FALSE))
```

#### Average Trip Duration By User Type 

```{r Average Trip Duration By User Type}
cyclistic_final_df %>% 
  group_by(user_type) %>% 
  summarize(avg_trip_duration = mean(trip_duration)) %>% 
  arrange(user_type) %>% 
  ggplot(aes(x = user_type, y = avg_trip_duration, fill = user_type)) +
  labs(title = "Average Trip Duration By User Type") +
  geom_col(width = 0.5, position = position_dodge(width = 0.5)) +
  scale_y_continuous(labels = function(x) format(x, scientific = FALSE))
```

#### Busiest Week By User Type

```{r Busiest Week By User Type}
cyclistic_final_df %>% 
  group_by(user_type, weekday) %>% 
  summarize(num_of_rides = n()) %>% 
  arrange(user_type, num_of_rides) %>% 
  ggplot(aes(x = weekday, y = num_of_rides, fill = user_type)) +
  labs(title = "Busiest Week Day By User Type") +
  geom_col(width = 0.5, position = position_dodge(width = 0.5)) +
  scale_y_continuous(labels = function(x) format(x, scientific = FALSE))
```

#### Total Monthly Rides By User Type

```{r Total Monthly Rides By User Type}
cyclistic_final_df %>%
  group_by(user_type, month)%>%
  summarize(num_of_rides = n())%>%
  arrange(user_type, month)%>%
  ggplot(aes(x = month, y = num_of_rides, fill = user_type)) +
  labs(title = "Total User Rides By Month") +
  geom_col(width = 0.5, position = position_dodge(width = 0.5)) +
  scale_y_continuous(labels = function(x) format(x, scientific = FALSE))
```

#### Most Popular Bike Type

```{r Most Popular Bike Type}
cyclistic_final_df %>% 
  group_by(bike_type, user_type) %>% 
  summarize(num_of_rides = n()) %>% 
  ggplot(aes(x = bike_type, y = num_of_rides, fill = user_type)) +
  labs(title = "Most Used Bike Type") +
  geom_col(width = 0.5, position = position_dodge(width = 0.5)) +
  scale_y_continuous(labels = function(x) format(x, scientific = FALSE))
```

#### Busiest Month By User Type

```{r Busiest Month By User Type}
cyclistic_final_df %>% 
  group_by(month, user_type) %>% 
  summarize(num_of_rides = n()) %>% 
  ggplot(aes(x = month, y = num_of_rides, fill = user_type)) +
  labs(title = "Number User Rides By Month") +
  geom_col(width = 0.5, position = position_dodge(width = 0.5)) +
  scale_y_continuous(labels = function(x) format(x, scientific = FALSE))
```

